---
title: "CEMA 2025 Data Science Internship"
author: "Emmanuel Rono"
date: "2025-04-28"
output: html_document
---

### Introduction

HIV remains one of the most pressing public health challenges in Africa, particularly in sub-Saharan regions, which bear the highest burden of the epidemic. Despite significant advancements in prevention, treatment, and awareness campaigns, HIV continues to affect millions of lives, with far-reaching consequences beyond healthcare, impacting economic stability, social equity, and overall development.

Tracking and analyzing HIV data is essential for understanding transmission trends, evaluating intervention effectiveness, and ensuring that healthcare systems are equipped to support those living with the virus. Additionally, persistent challenges such as stigma, limited access to care, and socioeconomic disparities underscore the need for data-driven strategies to guide policy decisions and resource allocation.

This markdown provides a comprehensive epidemiological analysis of HIV in Africa, examining key trends, regional disparities, and progress toward global targets. By leveraging data insights, we aim to inform targeted interventions, strengthen healthcare responses, and ultimately reduce the burden of HIV on individuals, communities, and health systems across the continent.

### Objectives

- Global Burden Analysis:
*"Develop an interactive time-series visualization highlighting HIV case trend in the countries that collectively account for 75% of the global disease burden."*

- Regional Burden Breakdown:
*"Create a visualization comparing HIV trends across WHO regions, focusing on the top countries contributing to 75% of cases within each region."*

#### Importing libraries

```{r}
# Load the packages 
library(tidyverse)
library(dplyr)
library(tidyverse)
library(plotly)
library(RColorBrewer)
library(purrr)
library(ggplot2)
library(tidyr)


# Read the HIV CSV file
hiv_df <- read_csv('C:/Users/Rono/Documents/internship_task_dscience/HIV data 2000-2023.csv')

# View the first few rows
head(hiv_df)
```
### Data Understanding

The hiv dataset contains the following columns with their descriptions:

-`IndicatorCode`: A unique identifier for the indicator. ie HIV_0000000001

-`Indicator`: A detailed description of what is being measured ie "Estimated number of people (all ages) living with HIV".

-`ValueType`: Identifies the data type whether numeric etc.

-`ParentLocationCode`: An abbreviation for the continent where a country is located in ie AFR for Africa.

-`ParentLocation`: Continent where a country is in (e.g., "Africa").

-`Location type`: Describes the type of location (e.g., "Country").

-`SpatialDimValueCode`: An abbreviation for specific country (e.g., "AGO" for Angola).

-`Location`: A specified country ie Angola

-`Period type`: time period (e.g., "Year").

-`Period`: The year for which the data is recorded (e.g., "2023").

-`Value`: The estimated number of people living with HIVgiven in a range (e.g., "320 000 [280 000 - 380 000]" for Angola in 2023).

This dataset provides detailed information on HIV across various countries and regions.


### Data Cleaning

```{r}
# Checking count of unique values in each column
for (column in colnames(hiv_df)) {
  unique_count <- length(unique(hiv_df[[column]]))
  cat(paste0("Number of unique values in '", column, "': ", unique_count, "\n"))
}
```

```{r}
# Remove columns with single values ie 1 unique value
hiv_df <- hiv_df[, !(names(hiv_df) %in% c("IndicatorCode", "Indicator", "ValueType", "Period type"))]
hiv_df
```
```{r}
# Check for null values
colSums(is.na(hiv_df))
```
```{r}
# Basic info about the dataframe
str(hiv_df)
```

We have no nulls. However in the Value column we can see some locations have 'No data' on people living with HIV.

```{r}
#checking for location with no data on HIV
locations_with_no_data <- unique(hiv_df$`Location`[hiv_df$Value == "No data"])
locations_with_no_data
```

```{r}

# filtering out locations with no data
hiv_df <- hiv_df %>% 
  filter(Value != "No data")
hiv_df
```
```{r}
# View the Value column
head(hiv_df$Value, 20)
```

The `value` column has a range and a an already specified figure. We also have scenarios where '<' is used.

```{r}
# synchronising the Value column
hiv_df <- hiv_df %>% 
  mutate(
    Value = case_when(
      str_detect(Value, "^<") ~ as.integer(str_extract(Value, "\\d+")) - 1L,
      TRUE ~ as.integer(str_replace_all(str_extract(Value, "^[\\d\\s]+"), " ", ""))
    )
  )

hiv_df
```

### Question 1a
##### Create a visualization that shows the trend of HIV cases in the countries that contribute to 75% of the global burden

```{r}
#create a summary table for each country and its yearly value
hiv_summary <- aggregate(Value ~ Location + Period, 
                        data = hiv_df, 
                        FUN = sum, 
                        na.rm = TRUE)
head(hiv_summary, 20)
```
```{r}


# Calculate yearly global cases
annual_hiv <- hiv_summary %>%
  group_by(Period) %>%
  summarise(Annual_Total = sum(Value, na.rm = TRUE)) %>%
  ungroup()

# View the result
annual_hiv
```
Now we need to find country %

- For each year, the global total of HIV cases is first calculated. This total is then paired with country-specific values for the same year, enabling the computation of each nation's annual percentage contribution to the global burden.


```{r}

# Merge and calculate percentages
merged_df <- hiv_summary %>%
  left_join(annual_hiv, by = "Period") %>%
  mutate(Percent = Value / Annual_Total * 100)

# View the result
head(merged_df, 20)
```

Our analysis shifts focus from *snapshot* averages to *cumulative burden* (2000–2023), identifying nations with the greatest sustained disease burden. This longitudinal approach prioritizes consistent contributors over transient spikes, revealing which countries have driven HIV prevalence persistently across two decades rather than temporarily.


```{r}
# Country totals
country_hiv <- aggregate(Value ~ Location, 
                        data = merged_df, 
                        FUN = sum, 
                        na.rm = TRUE)
names(country_hiv)[2] <- "Value"

# Global total
global_hiv_sum <- sum(annual_hiv$Annual_Total, na.rm = TRUE)
country_hiv
```

We then find the percentage each country contributed to this global total.

```{r}
# calculating %
country_hiv <- country_hiv %>%
  mutate(
    Percent = Value / global_hiv_sum * 100,
    Percent_Fmt = paste0(round(Percent, 2), "%")
  ) %>%
  select(Location, Value, Percent, Percent_Fmt) %>%
  arrange(desc(Percent))
country_hiv
```

Now we get the cumulative percentage.

```{r}
# Compute cumulative percent
country_hiv <- country_hiv %>%
  arrange(desc(Percent)) %>%
  mutate(Total_Percent = cumsum(Percent))

# Select countries contributing to 75% of total global cases
top_countries <- country_hiv %>%
  filter(Total_Percent <= 75) %>%
  pull(Location)  # Use `pull` to extract as a vector/list

# Print the result
top_countries
```

- Now we plot the line graghs for these countries in top %

```{r}

# create color palette
num_countries <- length(unique(top_countries))
custom_palette <- colorRampPalette(brewer.pal(8, "Set2"))(num_countries)

# Create the interactive plot
fig <- hiv_summary %>%
  # Filter for top countries contributing to 75% of cases
  filter(Location %in% top_countries) %>%
  
  # Create plotly visualization with custom colors
  plot_ly(
    x = ~Period,
    y = ~Value,
    color = ~Location,
    type = 'scatter',
    mode = 'lines+markers',
    colors = custom_palette,
    text = ~paste0(
      "<b>", Location, "</b><br>",
      "Year: ", Period, "<br>",
      "Cases: ", format(Value, big.mark = ",")
    ),
    hoverinfo = 'text'
  ) %>%
  
  # Add layout customization
  layout(
    title = list(
      text = "<b>Trend of HIV Cases in Top Contributing Countries</b><br><sub>(75% of Global Burden)</sub>",
      x = 0.05,
      xanchor = 'left'
    ),
    xaxis = list(title = "Year"),
    yaxis = list(title = "Estimated HIV Cases"),
    hovermode = "x unified",
    legend = list(orientation = 'h', y = -0.2),
    # Add range slider for time navigation
    xaxis = list(rangeslider = list(visible = TRUE))
  )

# Display the plot
fig
```

The interactive graph allows users to explore country-specific trends by clicking directly on the legend. Double-click any country’s name to view its data alone, or click multiple countries to compare them. To reset and see all countries together again, double-click any empty area in the legend. The tooltips display exact values when you hover over lines, and you can zoom into specific time periods or save the view using the toolbar buttons in the top-right corner.

### Question 1b
##### Generate a visualization that displays the trend of HIV cases in the countries contributing to 75% of the burden within each WHO region (column called ParentLocationCode contains the WHO regions)

- Who are the biggest players inside each WHO region?

```{r}
# grouping locations
who <- hiv_df %>%
  group_by(ParentLocationCode, Location, Period) %>%
  summarise(Total_Value = sum(Value, na.rm = TRUE)) %>%
  ungroup() 

who
```


To analyze regional HIV distributions, we first calculate annual totals for each WHO region by summing cases across all member countries. These regional aggregates then serve as the baseline to determine each country's percentage contribution within its respective geographic context, revealing relative burdens over time.


```{r}
# calculating regional totalss
regional_yearly_total <- who %>%
  group_by(ParentLocationCode, Period) %>%
  summarise(Regional_Total = sum(Total_Value, na.rm = TRUE)) %>%  # Sum values, ignoring nulls if any
  ungroup()  

# View the result
regional_yearly_total
```

We calculate annual percentage contributions by comparing each country's HIV case count against its WHO region's yearly total. This reveals how much of the regional burden each nation represents in a given year, allowing comparisons within geographic groups rather than globally.

```{r}
# comparison df
merged_df <- who %>%
  # Merge with regional totals
  left_join(regional_yearly_total, by = c("ParentLocationCode", "Period")) %>%
  # Calculate percentage contribution
  mutate(Percent = (Total_Value / Regional_Total) * 100)

# View the result
merged_df
```


We calculate cumulative HIV cases (2000–2023) for each country and region, quantifying their total historical contribution to the epidemic. This aggregates impact over decades, avoiding overemphasis on short-term fluctuations.


```{r}
# getting total contribution
total_contribution <- merged_df %>%
  group_by(ParentLocationCode, Location) %>%
  summarise(Value = sum(Total_Value, na.rm = TRUE)) %>%  # Sum across all years
  ungroup()  # Remove grouping

# View the result
total_contribution
```

Calculate cumulative regional totals by summing HIV cases across all years (2000–2023) for each WHO region, providing a comprehensive view of long-term disease burden distribution.

```{r}
#regional totals
regional_total_sum <- total_contribution %>%
  group_by(ParentLocationCode) %>%
  summarise(Regional_Sum = sum(Value, na.rm = TRUE)) %>%  # Sum per region
  ungroup()  # Remove grouping

# View the result
regional_total_sum
```

By integrating cumulative regional totals (2000–2023) with national-level data, we compute each country’s percentage contribution to its WHO region’s total HIV burden, revealing cross-temporal patterns in disease distribution.

```{r}
total_contribution <- total_contribution %>%
  # Merge with regional totals
  left_join(regional_total_sum, by = "ParentLocationCode") %>%
  # Calculate percentage contribution
  mutate(Percent = (Value / Regional_Sum) * 100)

# View the result
total_contribution
```

Within each WHO region, rank countries by their percentage contributions (descending order) and compute cumulative totals, revealing how concentrated the regional burden is among top contributors over time.

```{r}
total_contribution <- total_contribution %>%
  # Sort by region (ascending) and percent contribution (descending)
  arrange(ParentLocationCode, desc(Percent)) %>%
  # Group by region and calculate cumulative percentage
  group_by(ParentLocationCode) %>%
  mutate(CumulativePercent = cumsum(Percent)) %>%
  ungroup()  # Remove grouping for future operations

# View the result
total_contribution
```

Filter each WHO region to retain only the highest-ranked countries until their cumulative percentage contributions reach ≤75%, highlighting the subset of nations driving three-quarters of the region's total burden over the study period.

```{r}
#top countries/region
top_countries_per_region <- total_contribution %>%
  group_by(ParentLocationCode) %>%
  arrange(desc(Value)) %>%
  mutate(
    CumulativePercent = cumsum(Value)/sum(Value)*100
  ) %>%
  filter(CumulativePercent <= 75) %>%
  ungroup()
```



```{r}

# Merge ParentLocation names ----
region_names <- hiv_df %>%
  distinct(ParentLocationCode, ParentLocation)


top_countries_per_region <- top_countries_per_region %>%
  left_join(region_names, by = "ParentLocationCode")
```



```{r}

# Filter and prepare trend data ----
top_country_trend <- merged_df %>%
  filter(Location %in% top_countries_per_region$Location) %>%
  left_join(region_names, by = "ParentLocationCode") %>%
  arrange(ParentLocation, Period)

```


```{r}
# Create separate plots for each region ----
region_plots <- top_country_trend %>%
  group_split(ParentLocation) %>%
  map(~{
    region_name <- unique(.x$ParentLocation)
    
    # Generate color palette for countries in this region
    country_colors <- colorRampPalette(brewer.pal(8, "Set2"))(
      n_distinct(.x$Location)
    )
    
    plot_ly(.x, height = 500) %>%
      add_trace(
        x = ~Period,
        y = ~Total_Value,
        color = ~Location,
        colors = country_colors,
        type = "scatter",
        mode = "lines+markers",
        text = ~paste0(
          "<b>", Location, "</b><br>",
          "Year: ", Period, "<br>",
          "Cases: ", format(Total_Value, big.mark = ",")
        ),
        hoverinfo = "text"
      ) %>%
      layout(
        title = list(
          text = paste0("<b>", region_name, "</b><br>HIV Case Trends"),
          x = 0.05
        ),
        xaxis = list(
          title = "Year",
          rangeslider = list(visible = TRUE)
        ),
        yaxis = list(title = "Estimated Cases"),
        showlegend = TRUE,
        margin = list(b = 100, t = 100)
      )
  }) %>%
  setNames(unique(top_country_trend$ParentLocation))

# Render plots in HTML output ----
htmltools::tagList(region_plots)  # Directly use the plot list
```


This analysis visualizes HIV trends for the top contributing countries (covering ≤75% of regional cases) across WHO regions. For each region, it generates an interactive line chart showing yearly case numbers from 2000-2023. Countries are color-coded within their region, with hover tooltips displaying exact case counts and a slider to zoom into specific time periods. Charts are automatically split into separate views per region for clearer comparison of national trends within their geographic context.





### Question 1c
##### You have also been provided with World Bank data on the multidimensional poverty headcount ratio, which includes factors such as income, educational attainment, school enrolment, electricity access, sanitation and drinking water. We would like you to merge this dataset with the HIV data above and analyze the relationship between people living with HIV and multidimensional poverty, and the individual factors that contribute to the ratio. Remember to account for the random effects (country, year)

```{r}
# load dependencies
library(readxl)

# Read the file without headers
poverty_df <- read_excel("C:/Users/Rono/Documents/internship_task_dscience/multidimensional_poverty.xlsx", col_names = FALSE)

# Extract header rows (Excel rows 2 and 3)
header_row1 <- unlist(poverty_df[2, ])
header_row2 <- unlist(poverty_df[3, ])

# Combine headers
combined_headers <- mapply(function(h1, h2) {
  if (!is.na(h1) && !is.na(h2)) {
    paste(h1, h2, sep = "_")
  } else if (!is.na(h1)) {
    as.character(h1)
  } else {
    as.character(h2)
  }
}, header_row1, header_row2) %>% 
  make.names(unique = TRUE)

# Assign headers and clean up
names(poverty_df) <- combined_headers
poverty_df <- poverty_df %>%
  slice(-1, -2) %>%  # Remove header rows (original rows 2-3)
  tibble::rowid_to_column() %>%  # Reset index
  mutate(across(everything(), as.character))  # Convert all columns to character

# View result
poverty_df
```

```{r}
#check dtpes and info
str(poverty_df)
```

```{r}
# Rename all columns sequentially
colnames(poverty_df) <- c("RowID", 'Region', "Country Code", "Economy", 'Reporting Year', 'Survey Name', 'Survey Year', 'Survey Coverage', 'Welfare Type', 'Survey Comparability', 'Monetary (%)', 'Educational Attainment (%)', 'Educational Enrollment (%)', 'Electricity (%)', 'Sanitation (%)', 'Drinking Water (%)', 'Multidimensional Poverty Headcount Ratio (%)')
poverty_df
```


```{r}
# Drop first row
poverty_df <- poverty_df[-1, ]
poverty_df
```

```{r}
#check dtypes
str(poverty_df)
```


```{r}
# Convert column types ----
poverty_df <- poverty_df %>%
  mutate(
    # Integer conversions
    across(c("Reporting Year", "Survey Year", "Survey Comparability"),
           ~ as.integer(.)),
    
    # Numeric conversions (handling '-' as NA)
    across(c("Monetary (%)", "Educational Attainment (%)",
             "Educational Enrollment (%)", "Electricity (%)",
             "Sanitation (%)", "Drinking Water (%)",
             "Multidimensional Poverty Headcount Ratio (%)"),
           ~ as.numeric(na_if(., "-")))
  )
```

```{r}
#check dtypes
str(poverty_df)
```

```{r}
# Get missing values count per column
null_counts <- colSums(is.na(poverty_df))
print(null_counts)

```

```{r}
# Get columns with missing values
cols_with_na <- names(which(colSums(is.na(poverty_df)) > 0))

# Replace NAs with column medians
poverty_df <- poverty_df %>%
  mutate(across(all_of(cols_with_na), 
                ~ replace_na(., median(., na.rm = TRUE))))

# Custom info function to verify
df_info <- function(df) {
  cat("Data Types and Non-Null Counts:\n")
  data.frame(
    Column = names(df),
    Non_Null = sapply(df, function(x) sum(!is.na(x))),
    DType = sapply(df, function(x) class(x)[1])
  ) %>% print(row.names = FALSE)
}

# Check results
df_info(poverty_df)
```


```{r}
#drop columns
cols_to_drop <- c("RowID")
poverty_df <- subset(poverty_df, select = -which(names(poverty_df) %in% cols_to_drop))
poverty_df
```

#### Merge Hiv_df and Poverty_df

The HIV dataset and poverty dataset share common identifiers that enable cross-referencing:

- `SpatialDimValueCode` (HIV dataset) matches `Country code` (poverty dataset) for geographic alignment

- `Location` (HIV dataset) corresponds to `Economy` (poverty dataset) for country-level comparisons

- `Period` (HIV dataset) aligns with `Reporting year` (poverty dataset), both representing the actual calendar year of data collection

These equivalent fields facilitate temporal and spatial merging of health and socioeconomic indicators.


```{r}
# Merge the datasets on the common columns in R
merged_df <- merge(
  hiv_df,
  poverty_df,
  by.x = c("SpatialDimValueCode", "Location", "Period"),
  by.y = c("Country Code", "Economy", "Reporting Year"),
  all = FALSE  # equivalent to 'inner' join; use all = TRUE for 'outer' join
)

# View the merged dataframe
head(merged_df, 20)

```

Now to create graphs to see correlation between value and other columns


```{r}
# 

# Subset data and remove NA
corr_data <- merged_df %>%
  select(
    `Value`,  # HIV prevalence
    `Multidimensional Poverty Headcount Ratio (%)`,
    `Monetary (%)`,
    `Educational Attainment (%)`,
    `Educational Enrollment (%)`,
    `Electricity (%)`,
    `Sanitation (%)`,
    `Drinking Water (%)`
  ) %>%
  na.omit()

# Calculate and reshape the correlation matrix
cor_df <- cor(corr_data) %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "corr")

#  Create the ggplot heatmap
p <- ggplot(cor_df, aes(var1, var2, fill = corr, text = sprintf("Corr: %.2f", corr))) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", corr)), 
            color = "black", size = 4) +
  scale_fill_gradient2(low = "#0571b0", high = "#ca0020", mid = "white",
                       midpoint = 0, limit = c(-1, 1)) +
  labs(title = "Correlation Matrix: HIV Prevalence and Poverty Factors",
       x = "", y = "") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, face = "bold")) +
  coord_fixed()

#  Convert to interactive plot with zoom
ggplotly(p, tooltip = "text")
```
 The correlation plot can be zoomed for ease of vision.
    

#### **Conclusion**

##### **Insights**

1. Sanitation Infrastructure

Moderate association (r=0.46): Inadequate sanitation systems correlate with elevated HIV prevalence
Interpretation: Poor waste management infrastructure appears moderately linked to disease transmission pathways

2. Multidimensional Poverty Nexus

Strong positive correlation (r=0.73): Concurrent deprivations in health, education, and living standards show near-linear relationship with HIV rates
Policy Implication: Poverty alleviation programs should address intersecting deprivation factors3.

3. Water Access Crisis

Strongest correlation (r=0.74): Limited safe water access demonstrates particularly pronounced HIV association
Mechanism: Water scarcity may increase vulnerability through hygiene challenges and economic stressors

4. Economic Determinants

Near 1:1 relationship (r=0.80): Income poverty shows strongest connection to HIV prevalence
Critical Finding: Monetary deprivation appears fundamental in HIV transmission dynamics

5. Educational Disparities

Moderate linkage (r=0.40): Lower educational achievement correlates with increased HIV risk
Notable Gap: Education quality/type may be more impactful than mere attainment levels

6. School Enrollment Paradox

Negligible correlation (r=0.15): Educational access alone shows minimal HIV impact
Contrary Expectation: Enrollment rates don't significantly predict transmission patterns

7. Energy Poverty Connection

Substantial association (r=0.67): Electricity deprivation corresponds to heightened HIV prevalence
Development Priority: Energy infrastructure investments could yield health co-benefits


- The analysis reveals a poverty-HIV continuum where economic deprivation (r=0.80) and multidimensional poverty (r=0.73) emerge as primary drivers, while basic infrastructure deficits (water r=0.74, energy r=0.67) compound transmission risks. Educational factors show divergent impacts, suggesting need for targeted interventions rather than blanket education strategies.




#### Question 2

```{r}
# Read the CSV file into a dataframe
mortality_df <- read_csv("C:/Users/Rono/Documents/internship_task_dscience/dataset_datascience.csv")

# Display the dataframe
print(mortality_df)
```

```{r}
#load libararies
library(scales)    
library(sf)      
```


```{r}
#view columns
colnames(mortality_df)
```

#### Filter data for the eight countries belonging to the East African Community.

```{r}
# Define EAC countries
eac_countries <- c("Burundi", "Democratic Republic of the Congo", "Kenya", 
                  "Rwanda", "Somalia", "South Sudan", "Tanzania", "Uganda")

# Filter mortality data for EAC countries
eac_mortality <- mortality_df %>%
  filter(`Geographic area` %in% eac_countries)

# View filtered data
print(eac_mortality)
```


```{r}
#check info
glimpse(eac_mortality)
```
```{r}
# dropping columns

eac_mortality_dropped <- eac_mortality %>%
  select(-`Regional group`, -`Country notes`, -Definition)

# View the modified dataframe
print(eac_mortality_dropped)
```


```{r}
# Get value counts with percentages
eac_mortality_dropped %>%
  count(`Geographic area`, sort = TRUE) %>%
  mutate(percent = n / sum(n) * 100)
```

We dont have data on Tanzania


#### Visualize the latest estimate of each indicator at the country level using shapefiles

```{r}
# EAC country codes
eac_codes <- c("BDI", # Burundi
               "KEN", # Kenya
               "RWA", # Rwanda
               "SSD", # South Sudan
               "UGA", # Uganda
               "COD", # Democratic Republic of the Congo
               "SOM") # Somalia
```


```{r}
# Load necessary libraries
library(sf)
library(purrr)
library(dplyr)

# Define EAC ISO codes (GADM uses ISO3 codes)
eac_codes <- c("BDI", "COD", "KEN", "RWA", "SOM", "SSD", "TZA", "UGA")

# Create a folder to store the downloaded shapefiles
folder_path <- "eac_shapefiles"
if (!dir.exists(folder_path)) dir.create(folder_path)

# Function to download and load shapefiles for one country
load_country_shape <- function(iso) {
  url <- paste0("https://geodata.ucdavis.edu/gadm/gadm4.1/shp/gadm41_", iso, "_shp.zip")
  temp <- tempfile(fileext = ".zip")
  
  message("Downloading shapefile for ", iso, "...")
  tryCatch({
    download.file(url, temp, mode = "wb", quiet = TRUE)
    unzip(temp, exdir = file.path(folder_path, iso))
    shape_path <- file.path(folder_path, iso, paste0("gadm41_", iso, "_0.shp"))
    st_read(shape_path, quiet = TRUE)
  }, error = function(e) {
    message("Failed to load ", iso, ": ", e$message)
    return(NULL)
  })
}

# Apply the function and merge all valid shapefiles
eac_shapes <- map(eac_codes, load_country_shape) %>%
  compact() %>%  # remove NULLs in case of failed downloads
  bind_rows()

# View summary
print(eac_shapes)
plot(st_geometry(eac_shapes), main = "EAC Countries Map")

```
```{r}
# Combine all individual country shapefiles into one sf (GeoDataFrame equivalent)
eac_gdf <- do.call(rbind, eac_shapes)

# View summary of the combined spatial data
print(eac_gdf)

# Remove NULLs and bind all into one sf data frame
eac_gdf <- eac_shapes %>%
  purrr::compact() %>%   # removes NULLs
  dplyr::bind_rows()     # binds into a single sf object

# Confirm it's an sf object
class(eac_gdf)  # should include "sf"

# Plot the geometry
plot(st_geometry(eac_gdf), main = "Combined EAC Countries")
```

```{r}
# Filter for rows with the latest year in Series Year
final_yr <- eac_mortality_dropped %>%
  filter(`Series Year` == max(`Series Year`, na.rm = TRUE))

# View the result
print(final_yr)

```

```{r}
# Perform the left join using the correct keys
eac_merge <- eac_gdf %>%
  left_join(final_yr, by = c("GID_0" = "REF_AREA"))

# View the merged spatial dataframe
print(eac_merge)

```

```{r}
# Load the necessary package
library(dplyr)

# Filter for rows where the Indicator column equals "Neonatal mortality rate"
neonatal_df <- eac_merge %>%
  filter(Indicator == "Neonatal mortality rate")

# View the result
print(neonatal_df)
```

```{r}
# Load required packages
library(dplyr)
library(ggplot2)
library(sf)
library(readr)
library(tidyr)
library(rnaturalearth)
library(rnaturalearthdata)
library(viridis)

# ---------------- Load and Prepare Data ----------------
# Read the mortality dataset
mortality_data <- read_csv("C:/Users/Rono/Documents/internship_task_dscience/dataset_datascience.csv")

# Filter for East African Community countries
eac_countries <- c("Burundi", "Democratic Republic of the Congo", "Kenya", 
                   "Rwanda", "Somalia", "South Sudan", "Tanzania", "Uganda")
eac_data <- mortality_data %>%
  filter(`Geographic area` %in% eac_countries)

# Clean the data
eac_data <- eac_data %>%
  filter(!is.na(`Observation Value`), `Observation Value` != "..") %>%
  mutate(`Observation Value` = as.numeric(`Observation Value`))

# Use only the latest year
latest_year <- max(eac_data$`Series Year`, na.rm = TRUE)
eac_latest <- eac_data %>%
  filter(`Series Year` == latest_year)

# ---------------- Get World Map Data ----------------
# Use Natural Earth low-resolution country boundaries
world <- ne_countries(scale = "medium", returnclass = "sf")

# Filter for EAC countries
eac_map <- world %>% filter(name %in% eac_countries)

# ---------------- Merge Neonatal Mortality ----------------
neonatal_data <- eac_latest %>%
  filter(Indicator == "Neonatal mortality rate") %>%
  group_by(`Geographic area`) %>%
  summarise(NMR = mean(`Observation Value`, na.rm = TRUE))

eac_neonatal_map <- left_join(eac_map, neonatal_data, by = c("name" = "Geographic area"))

# ---------------- Plot Neonatal Mortality ----------------
ggplot(eac_neonatal_map) +
  geom_sf(aes(fill = NMR), color = "black") +
  scale_fill_viridis(option = "inferno", na.value = "grey90", name = "NMR") +
  labs(
    title = "Neonatal Mortality Rate (per 1,000 live births)",
    subtitle = paste("Latest Year:", latest_year),
    caption = "Data Source: UNICEF / UN Stats"
  ) +
  theme_minimal()

```



```{r}
# ---------------- Load and Prepare Data ----------------
# Read the mortality dataset
mortality_data <- read_csv("C:/Users/Rono/Documents/internship_task_dscience/dataset_datascience.csv")

# Filter for East African Community countries
eac_countries <- c("Burundi", "Democratic Republic of the Congo", "Kenya", 
                   "Rwanda", "Somalia", "South Sudan", "Tanzania", "Uganda")
eac_data <- mortality_data %>%
  filter(`Geographic area` %in% eac_countries)

# Clean the data
eac_data <- eac_data %>%
  filter(!is.na(`Observation Value`), `Observation Value` != "..") %>%
  mutate(`Observation Value` = as.numeric(`Observation Value`))

# Use only the latest year
latest_year <- max(eac_data$`Series Year`, na.rm = TRUE)
eac_latest <- eac_data %>%
  filter(`Series Year` == latest_year)

# ---------------- Get World Map Data ----------------
# Use Natural Earth low-resolution country boundaries
world <- ne_countries(scale = "medium", returnclass = "sf")

# Filter for EAC countries
eac_map <- world %>% filter(name %in% eac_countries)

# ---------------- Merge Under-5 Mortality ----------------
under5_data <- eac_latest %>%
  filter(Indicator == "Under-five mortality rate") %>%
  group_by(`Geographic area`) %>%
  summarise(U5MR = mean(`Observation Value`, na.rm = TRUE))

eac_under5_map <- left_join(eac_map, under5_data, by = c("name" = "Geographic area"))

# ---------------- Plot Under-5 Mortality ----------------
ggplot(eac_under5_map) +
  geom_sf(aes(fill = U5MR), color = "black") +
  scale_fill_viridis(option = "inferno", na.value = "grey90", name = "U5MR") +
  labs(
    title = "Under-Five Mortality Rate (per 1,000 live births)",
    subtitle = paste("Latest Year:", latest_year),
    caption = "Data Source: UNICEF / UN Stats"
  ) +
  theme_minimal()

```


#### Show the average trends in the mortality rates over time (plot the average trend line and add the points in the graphic for the country level estimates for each indicator. Expectation: two plots)

```{r}
# Load libraries
library(tidyverse)
library(readr)
library(ggplot2)

# Load dataset
mortality_data <- read_csv("C:/Users/Rono/Documents/internship_task_dscience/dataset_datascience.csv")

# Filter for East African Community countries
eac_countries <- c("Burundi", "Democratic Republic of the Congo", "Kenya", 
                   "Rwanda", "Somalia", "South Sudan", "Tanzania", "Uganda")

eac_data <- mortality_data %>%
  filter(`Geographic area` %in% eac_countries) %>%
  filter(!is.na(`Observation Value`), `Observation Value` != "..") %>%
  mutate(`Observation Value` = as.numeric(`Observation Value`))

# Average multiple observations per country, year, and indicator
eac_avg <- eac_data %>%
  group_by(`Geographic area`, `Series Year`, Indicator) %>%
  summarise(mean_value = mean(`Observation Value`, na.rm = TRUE), .groups = "drop")

# Pivot data to get Under-5 and Neonatal in separate columns
eac_pivot <- eac_avg %>%
  pivot_wider(names_from = Indicator, values_from = mean_value)

# Drop rows with missing values in either column
eac_pivot <- eac_pivot %>%
  drop_na(`Under-five mortality rate`, `Neonatal mortality rate`)

# Plot Under-5 Mortality Rate
ggplot(eac_pivot, aes(x = `Series Year`, y = `Under-five mortality rate`, color = `Geographic area`)) +
  geom_point(alpha = 0.4) +
  stat_summary(fun = mean, geom = "line", aes(group = 1), color = "black", size = 1.2) +
  labs(
    title = "Average Trend of Under-5 Mortality Rate Over Time (East Africa)",
    x = "Year", y = "Under-five mortality rate (per 1,000 live births)",
    color = "Country"
  ) +
  theme_minimal() +
  theme(legend.position = "right") +
  guides(color = guide_legend(override.aes = list(size = 3)))

# Plot Neonatal Mortality Rate
ggplot(eac_pivot, aes(x = `Series Year`, y = `Neonatal mortality rate`, color = `Geographic area`)) +
  geom_point(alpha = 0.4) +
  stat_summary(fun = mean, geom = "line", aes(group = 1), color = "black", size = 1.2) +
  labs(
    title = "Average Trend of Neonatal Mortality Rate Over Time (East Africa)",
    x = "Year", y = "Neonatal mortality rate (per 1,000 live births)",
    color = "Country"
  ) +
  theme_minimal() +
  theme(legend.position = "right") +
  guides(color = guide_legend(override.aes = list(size = 3)))

```



Somalia and South Sudan has the highest estimate for the Under-five mortality rate.

Somalia, Rwanda and South Sudan show the highest record of neonatal mortality rate.


